---
title: "Comparative Survival Analysis of Drug A vs Drug B Using Simulated Real‑World Data"
output:
  github_document:
    toc: true
    toc_depth: 2
---

# Executive summary

This report demonstrates a **comparative effectiveness** workflow for observational oncology research using simulated registry‑like data:

- Describe baseline differences between treatment groups (**confounding check**).
- Compare crude survival using Kaplan–Meier curves (**descriptive**).
- Estimate an adjusted association using a multivariable Cox model (**measured confounding control**).
- Re‑estimate using **propensity score inverse probability of treatment weighting (IPTW)** as a robustness analysis.
- Explore **effect modification** by biomarker subgroup.

# Background and objective

In real‑world oncology research, treatment allocation is not randomized. Patients receiving different therapies often differ in age, disease severity, comorbidity, and socioeconomic factors, which can confound treatment–outcome associations.

**Primary objective:** estimate the association between **Drug A vs Drug B** and **overall survival** (time to death).

**Secondary objectives:**
- Assess baseline imbalances (standardized mean differences, SMD).
- Compare crude survival curves (Kaplan–Meier + log‑rank).
- Evaluate robustness using propensity score weighting (stabilized IPTW).
- Explore heterogeneity of treatment effect by biomarker (A/B/C).

# Data description

Simulated cohort (target N ≈ 12,000 patients), intended to mimic linked register data.

- **Exposure:** `treat_drugA` (0 = Drug B, 1 = Drug A)
- **Outcome:** `time_years` (follow‑up time in years), `event_death` (1 = death, 0 = censored)
- **Baseline covariates (measured):**
  - Demographics: `age`, `sex_male`
  - Disease severity: `stage_IV`
  - Health status: `comorbidity_score`, `prior_hosp_days_1y`
  - Socioeconomic: `ses` (Low/Medium/High)
  - Geography / health system: `region`
  - Calendar time: `year`
  - Biology: `biomarker` (A/B/C)

# Methods

## Overview

1. **Baseline comparison** by treatment group  
   - Create a Table 1 and compute SMD (imbalance threshold: |SMD| > 0.10).

2. **Kaplan–Meier (KM)**  
   - Plot unadjusted survival curves, report 1/3/5‑year survival, and log‑rank p‑value.

3. **Adjusted Cox proportional hazards model**  
   - Fit: `Surv(time_years, event_death) ~ treat + covariates`
   - Report adjusted hazard ratio (HR) for Drug A vs Drug B with 95% CI.
   - Check proportional hazards (PH) using `cox.zph()`.

4. **Propensity score IPTW (robustness)**  
   - Estimate propensity score (PS): `P(Drug A | covariates)`
   - Compute **stabilized weights** to estimate an **ATE** (average treatment effect).
   - Assess post‑weighting balance (SMD + love plot).
   - Fit weighted Cox model with robust SE.
   - Optional: weight truncation at 1st/99th percentiles to reduce influence of extreme weights.

5. **Effect modification**  
   - Fit interaction model: `treat * biomarker` + confounders
   - Derive HR within each biomarker group via linear combinations.
   - Visualize subgroup HRs (forest plot).

---

```{r setup }
# Global chunk options for a clean report
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)

# Load packages
required_pkgs <- c(
  "dplyr", "ggplot2",
  "survival", "survminer",
  "tableone", "broom",
  "cobalt"
)

missing_pkgs <- setdiff(required_pkgs, rownames(installed.packages()))
if (length(missing_pkgs) > 0) {
  stop("Missing packages: ", paste(missing_pkgs, collapse = ", "),
       ". Please install them, e.g. install.packages(c(",
       paste(sprintf('"%s"', missing_pkgs), collapse = ", "),
       ")).")
}

library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(tableone)
library(broom)
library(cobalt)
```

# 1. Import and prepare data

```{r import}
data_raw <- read.csv("pharmacoepi_survival_simulated.csv", stringsAsFactors = FALSE)

# Quick peek
dplyr::glimpse(data_raw)
head(data_raw)
```

```{r clean}
# Data cleaning + type casting
# - Keep original variables, but create human-readable factors for modeling/plots
data <- data_raw %>%
  mutate(
    # treatment coding: 0/1 -> factor with labels
    treat_drugA = as.integer(treat_drugA),
    treat = factor(treat_drugA, levels = c(0, 1), labels = c("Drug B", "Drug A")),

    # outcome/event
    event_death = as.integer(event_death),

    # covariates
    age = as.numeric(age),
    sex_male = as.integer(sex_male),
    stage_IV = factor(stage_IV, levels = c(0, 1), labels = c("No", "Yes")),
    comorbidity_score = as.integer(comorbidity_score),
    prior_hosp_days_1y = as.integer(prior_hosp_days_1y),
    ses = factor(ses, levels = c("Low", "Medium", "High")),
    region = factor(region),
    year = as.integer(year),
    biomarker = factor(biomarker, levels = c("A", "B", "C"))
  )

# Missingness table (counts of NA by variable)
missing_tbl <- data %>% summarise(across(everything(), ~sum(is.na(.))))
missing_tbl
```

> **Analysis dataset:** For simplicity, models below use complete‑case rows for the variables in each model.  
> In applied work, you would consider multiple imputation or other missing data strategies when appropriate.

# 2. Baseline characteristics 

```{r table1}
# Variables for baseline comparison
vars <- c(
  "age", "sex_male", "stage_IV",
  "comorbidity_score", "prior_hosp_days_1y",
  "ses", "region", "year", "biomarker"
)

tab1 <- CreateTableOne(
  vars = vars,
  strata = "treat",
  data = data,
  test = FALSE
)

# Print Table 1 with standardized mean differences
print(tab1, smd = TRUE)
```

```{r smd_top}
# Extract SMD and list the largest imbalances (absolute SMD)
smd_vec <- ExtractSmd(tab1)
smd_tbl <- tibble::tibble(
  variable = names(smd_vec),
  smd = as.numeric(smd_vec),
  abs_smd = abs(as.numeric(smd_vec))
) %>%
  arrange(desc(abs_smd))

smd_tbl %>% head(10)
```

# 3. Unadjusted survival (Kaplan–Meier)

```{r km}
# KM fit
df_km <- data %>%
  select(time_years, event_death, treat) %>%
  filter(complete.cases(.))

fit_km <- survfit(Surv(time_years, event_death) ~ treat, data = df_km)

# Survival at 1, 3, 5 years
km_times <- c(1, 3, 5)
km_summ <- summary(fit_km, times = km_times)

# Tidy survival estimates into a table
km_tbl <- data.frame(
  time_years = km_summ$time,
  treat = gsub("^treat=", "", km_summ$strata),
  surv = km_summ$surv,
  lower = km_summ$lower,
  upper = km_summ$upper
)

km_tbl
```

```{r km_plot}
# Log-rank test (unadjusted)
lr <- survdiff(Surv(time_years, event_death) ~ treat, data = df_km)
logrank_p <- 1 - pchisq(lr$chisq, df = length(lr$n) - 1)

# Plot KM curves
ggsurvplot(
  fit_km,
  data = df_km,
  risk.table = TRUE,
  risk.table.height = 0.30,
  risk.table.fontsize = 3.2,
  conf.int = TRUE,
  pval = TRUE,
  pval.method = TRUE,
  pval.size = 5,
  xlab = "Years since treatment start",
  ylab = "Overall survival probability",
  legend.title = "Treatment",
  legend.labs = c("Drug B", "Drug A")
)
```

# 4. Adjusted Cox model (primary)

```{r cox_primary}
# Primary adjusted Cox PH model
m1 <- coxph(
  Surv(time_years, event_death) ~
    treat + age + sex_male + stage_IV +
    comorbidity_score + prior_hosp_days_1y +
    ses + region + year + biomarker,
  data = data
)

summary(m1)

# Tidy HR table (exponentiated coefficients)
cox_table <- tidy(m1, exponentiate = TRUE, conf.int = TRUE)
cox_table
```

```{r cox_primary_hr}
# Pull the main treatment effect (Drug A vs Drug B)
hr_primary <- cox_table %>%
  filter(term == "treatDrug A") %>%
  transmute(
    HR = estimate,
    LCL = conf.low,
    UCL = conf.high,
    p_value = p.value,
    HR_CI = sprintf("%.2f (%.2f–%.2f)", HR, LCL, UCL)
  )

hr_primary
```

```{r ph_assumption}
# PH assumption check (Schoenfeld residual test)
ph_test <- cox.zph(m1)
ph_test
```

# 5. Propensity score IPTW (robustness)

## 5.1 Estimate propensity scores

```{r ps_model}
# Propensity score model: P(Drug A | covariates)
# Note: For stability, keep the same covariates as in the outcome model where possible.
ps_mod <- glm(
  I(treat == "Drug A") ~
    age + sex_male + stage_IV +
    comorbidity_score + prior_hosp_days_1y +
    ses + region + year + biomarker,
  data = data,
  family = binomial()
)

data_w <- data %>%
  mutate(
    ps = predict(ps_mod, type = "response"),
    treat_ind = as.integer(treat == "Drug A")  # 1=A, 0=B
  )

summary(data_w$ps)
```

```{r ps_overlap}
# Propensity score overlap (positivity diagnostic)
ggplot(data_w, aes(x = ps, fill = treat)) +
  geom_histogram(bins = 40, position = "identity", alpha = 0.4) +
  labs(
    fill = "Treatment",
    x = "Propensity score",
    y = "Count",
    title = "Propensity score distribution by treatment"
  )
```

## 5.2 Stabilized weights + diagnostics

```{r iptw_weights}
# Stabilized IPTW for ATE
p_treat <- mean(data_w$treat_ind == 1, na.rm = TRUE)
p_ctrl  <- 1 - p_treat

data_w <- data_w %>%
  mutate(
    sw = ifelse(treat_ind == 1, p_treat / ps, p_ctrl / (1 - ps))
  )

summary(data_w$sw)

# Optional truncation to reduce influence of extreme weights
w_lo <- quantile(data_w$sw, 0.01, na.rm = TRUE)
w_hi <- quantile(data_w$sw, 0.99, na.rm = TRUE)

data_w <- data_w %>%
  mutate(sw_trunc = pmin(pmax(sw, w_lo), w_hi))

summary(data_w$sw_trunc)
```

```{r iptw_weight_plot}
# Weight distribution (truncated view for readability)
ggplot(data_w, aes(x = sw_trunc)) +
  geom_histogram(bins = 60) +
  labs(x = "Stabilized weights (truncated at 1st/99th pct)", y = "Count",
       title = "Distribution of stabilized IPTW")
```

## 5.3 Balance assessment (before vs after weighting)

```{r balance}
# Balance using cobalt after IPTW
# Tip: set `un = TRUE` to get BOTH unweighted and weighted balance in one object
bal_iptw <- bal.tab(
  treat_ind ~ age + sex_male + stage_IV + comorbidity_score + prior_hosp_days_1y +
    ses + region + year + biomarker,
  data = data_w,
  weights = data_w$sw_trunc,
  estimand = "ATE",
  un = TRUE
)

# Print balance summary (includes Unadjusted + Adjusted SMD)
bal_iptw
```

```{r loveplot}
# Love plot (visual check of balance)
love.plot(
  bal_iptw,
  stat = "mean.diffs",
  abs = TRUE,
  thresholds = c(m = 0.1),
  var.order = "unadjusted",
  sample.names = c("Unweighted", "Weighted (IPTW)"),
  title = "Covariate balance (|SMD|) before vs after IPTW"
)
```

## 5.4 Weighted Cox model

```{r cox_iptw}
# Weighted Cox model using stabilized, truncated weights
m_iptw <- coxph(
  Surv(time_years, event_death) ~ treat,
  data = data_w,
  weights = sw_trunc,
  robust = TRUE
)

summary(m_iptw)

iptw_table <- tidy(m_iptw, exponentiate = TRUE, conf.int = TRUE)
iptw_table
```

```{r compare_models}
# Compare treatment HR across approaches
hr_iptw <- iptw_table %>%
  filter(term == "treatDrug A") %>%
  transmute(
    HR = estimate,
    LCL = conf.low,
    UCL = conf.high,
    p_value = p.value,
    HR_CI = sprintf("%.2f (%.2f–%.2f)", HR, LCL, UCL)
  )

bind_rows(
  Primary_Adjusted_Cox = hr_primary,
  IPTW_Weighted_Cox = hr_iptw,
  .id = "model"
) %>% select(model, HR_CI, p_value)
```

# 6. Effect modification by biomarker

```{r interaction_model}
# Interaction model: treat * biomarker
m_int <- coxph(
  Surv(time_years, event_death) ~
    treat * biomarker +
    age + sex_male + stage_IV +
    comorbidity_score + prior_hosp_days_1y +
    ses + region + year,
  data = data
)

summary(m_int)
```

```{r subgroup_hr}
# Compute subgroup HRs for Drug A vs Drug B within each biomarker group
b <- coef(m_int)
V <- vcov(m_int)

hr_lincomb <- function(L, b, V) {
  est <- sum(L * b)
  se  <- sqrt(t(L) %*% V %*% L)
  data.frame(
    HR  = exp(est),
    LCL = exp(est - 1.96 * se),
    UCL = exp(est + 1.96 * se)
  )
}

# Name vector length
L_base <- rep(0, length(b)); names(L_base) <- names(b)

# Reference biomarker is "A" by factor level order above
L_A <- L_base; L_A["treatDrug A"] <- 1

L_B <- L_base; L_B["treatDrug A"] <- 1
if ("treatDrug A:biomarkerB" %in% names(b)) L_B["treatDrug A:biomarkerB"] <- 1

L_C <- L_base; L_C["treatDrug A"] <- 1
if ("treatDrug A:biomarkerC" %in% names(b)) L_C["treatDrug A:biomarkerC"] <- 1

subg <- dplyr::bind_rows(
  cbind(biomarker = "A", hr_lincomb(L_A, b, V)),
  cbind(biomarker = "B", hr_lincomb(L_B, b, V)),
  cbind(biomarker = "C", hr_lincomb(L_C, b, V))
)

subg
```

```{r forest}
# Forest plot for subgroup HRs
ggplot(subg, aes(x = biomarker, y = HR, ymin = LCL, ymax = UCL)) +
  geom_pointrange() +
  geom_hline(yintercept = 1, linetype = "dashed") +
  coord_flip() +
  labs(
    x = "Biomarker group",
    y = "Hazard Ratio (Drug A vs Drug B)",
    title = "Treatment effect by biomarker subgroup (interaction model)"
  )
```

# 7. Results (auto-populated when knitted)

## Baseline imbalance

Key imbalances (top 5 by |SMD|):

```{r results_smd, echo=FALSE}
smd_tbl %>% head(5)
```

Interpretation:
- |SMD| > 0.10 suggests potentially meaningful imbalance, which motivates adjusted analyses.

## Crude survival

Log‑rank p‑value (Drug A vs Drug B): **`r signif(logrank_p, 3)`**

1/3/5‑year KM survival estimates:

```{r results_km, echo=FALSE}
km_tbl
```

## Adjusted association (primary Cox)

Adjusted HR for Drug A vs Drug B: **`r hr_primary$HR_CI`** (p = `r signif(hr_primary$p_value, 3)`)

## IPTW robustness

IPTW‑weighted HR for Drug A vs Drug B: **`r hr_iptw$HR_CI`** (p = `r signif(hr_iptw$p_value, 3)`)

Balance after weighting:
- Maximum |SMD| after weighting: **`r round(max(abs(bal_iptw$Balance$Diff.Adj), na.rm=TRUE), 3)`**  
  (target: < 0.10)

## Biomarker heterogeneity

Subgroup HRs from interaction model:

```{r results_subg, echo=FALSE}
subg %>% mutate(
  HR_CI = sprintf("%.2f (%.2f–%.2f)", HR, LCL, UCL)
) %>% select(biomarker, HR_CI)
```

# 8. Discussion

## Interpretation

- The crude KM comparison reflects **both treatment effect and confounding**.
- The adjusted Cox model estimates the association after controlling for measured confounders.
- IPTW creates a pseudo‑population with improved covariate balance, reducing model dependence.
- Biomarker interaction explores whether the treatment association differs across biological subtypes.

## Assumptions and limitations

- **No unmeasured confounding:** as in any observational analysis, residual confounding may remain.
- **Positivity (overlap):** limited PS overlap can make IPTW unstable; inspect PS plots and weight distributions.
- **PH assumption:** evaluate Schoenfeld tests/plots; consider time‑varying effects if violated.
- **Simulated data:** results are illustrative and should not be interpreted as clinical truth.
